Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-05 pharmaco_nejm.Rmd
Fixed to drop controls no longer current

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf")

@@

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.  This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in drugs in almost all chapters were associated with elevated rate ratios for COVID-19.  We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by co-morbidity or frailty, causal effects of drugs on disease susceptibility could not be ruled out.

Thus the primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year. Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.  A further aim was to determine whether the effects underlying this association are sparse or diffuse.  In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.  In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive [and had not embarked -- to be checked in the CHI database by SW]. Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  BNF subparagraph codes generally represent pharmacological drug classes.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Otherf".  Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.

# Statistical analysis 
Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.  As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  ICD-10 and BNF codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565) For this analysis this list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an dispensed prescription of a drug for which that condition is the only indication.  Diagnoses of diabetes were extracted from linkage to the national diabetes register and supplemented by ICD-10 codes for diabetes and BNF codes for diabetes drugs for recently diagnosed persons not on the last extract of the register .  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material. <!-- clone the github repo -->

To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we carried out a backward stepwise variable selection procedure. 
Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.  The model with the highest log-likelihood was used to select which drug class should be dropped.  This procedure was repeated with the retained variables until only a single drug class remained.

To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, an independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls. By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.  The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription. `r nrow(coeffs.propensity)` drug classes were retained in the propensity score.  As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.

# Results
## Relation of severe COVID-19 to polypharmacy 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive). Table \ref{tab:numdrugsgr} shows that risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over. The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).  In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.

To study further the relationship with polypharmacy, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.  The rationale for this restriction was to reduce the frequency of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and COVID-19.  In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.  Supplementary Table \ref{tab:numdrugs.listed.any} shows that the association with COVID-19 was stronger in those without a diagnosis of a listed condition than in those with one or more listed conditions.  Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped. Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.  Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.  The last drug class retained by backward selection, and thus the variable with largest independent explanatory power for the effect of polypharmacy, was proton pump inhibitors.

## Associations of severe COVID-19 with proton pump inhibitor use
As the association of COVID-19 and exposure to proton pump inhibitors is not obviously explained by established co-morbidities, we explored this association in more depth.  As this examined a specific exposure, rather than polypharmacy which may have different profiles in care home residents or those diagnosed with listed conditions, the observations were not restricted to exclude care home residents or those diagnosed with a listed condition. Distinguishing between causality and confounding is challenging because proton pump inhibitors are usually co-prescribed with other drugs.  We used several approaches:

@@

* Adjusting for prespecified covariates: care home residence, SIMD quintile, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.

@@

Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].  cy.

Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section. The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Supplementary Table \ref{tab:propensity}.  As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor. Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.

## Associations with other drug classes
Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).  These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.

Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.  Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.

Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.  Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and ACE inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis. The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.  Without more data on health status from primary care records, the confounders cannot be accurately measured.  We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.  
The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].

Although a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy, this procedure is likely to retain the drug classes that are most commonly prescribed.  We have shown a clear dose-response relationship of dispensed average daily dose of proton pump inhibitors to risk of severe COVID-19.  Adjustment for covariates pre-specified as likely to confound this association -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly. The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.

The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.  Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.  In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small. One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.  We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.  However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.

It is thus not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.  Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.  If the effect is causal, the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.

As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through more persistent effects on the gut microbiome.  Uncertainties about the effect of proton pump inhibitors on the risk of community acquired pneumonia [@othman_community_2016] are not necessarily relevant to effects on enteric infection.  Proton pump inhibitors have been reported to affect the gut microbiome [@imhann_proton_2016].  If effects of proton pump inhibitor use on susceptiblity to COVID-19 are mediated through alteration of the gut microbiome, these effects would not necessarily be detected in a comparison of time windows. 
 <!---barrier to viral infection, or through altering the microbiome allowing with Warnings been issued about increased risk of _Cl. difficile_ infection from use proton pump inhibitors.  [ https://bnf.nice.org.uk/drug/omeprazole.html#importantSafetyInformations ] Given this possible association with increased risk of enteric infections --->.

>>>>

# rmarkdown::render("pharmaco_nejm.Rmd", output_file="pharmaco.pdf")

@@

# Background
The REACT-SCOT study is a matched case-control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.  This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  The objective of this study was to investigate possible explanations for this association: confounding by co-morbidity  factors, or direct effects of prescribed drugs.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test.  For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive. For this study based on ascertainment of test-positive , up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2. The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology and used to calculate average daily doses were calculated for each individual as the sum of (strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).  As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  ICD-10 diagnostic and BNF drug codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).  For this analysis this list was extended to include non-cardiac circulatory disease as initial analysis of this case-control study had shown this to be a risk factor for COVID-19.  Diagnoses of diabetes were extracted from linkage to the national diabetes register.

# Statistical analysis 
Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`. To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure. Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.  The model with the highest log-likelihood was used to select which drug class should be dropped.  This procedure was repeated with the retained variables until only a single drug class remained.

To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, an independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls. The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription. `r nrow(coeffs.propensity)` drug classes were retained in the propensity score.  As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.

# Results
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the observation period. Table \ref{tab:numdrugsgr} shows that risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over. The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).  In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.

We next investigated whether specific drug classes might underlie this association with polypharmacy.  To reduce confounding by multimorbidity and frailty, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.  Supplementary Table \ref{tab:numdrugs.listed.any} shows that the association with COVID-19 was stronger in those without a diagnosis of a listed condition than in those with one or more listed conditions.  We examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  The drug classes dropped at this stage included some of the most widely-prescribed drug classes that were associated with COVID-19 in univariate analyses, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The `r length(rows.print)` drug classes retained at this stage included loop diuretics, calcium channel blockers, and non-opioid analgesics.  The last drug class retained by backward selection, and thus the variable with largest independent explanatory power for the effect of polypharmacy, was proton pump inhibitors.

As the association with proton pump inhibitors was not obviously attributable to confounding by conditions designated as risk conditions for COVID-19, we explored it in more depth.  For these analyses the dataset was not restricted to exclude care home residents or those diagnosed with a listed condition as we were examining the effects of a specific drug class rather than polypharmacy.  We used several approaches to distinguish between causality and confounding:

@@

* Adjusting for prespecified covariates: care home residence, quintile of index of socioeconomic deprivation, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.

@@

Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].  Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section. The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Supplementary Table \ref{tab:propensity}.  As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor. Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained only in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.

## Associations with other drug classes
Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal), selected  because drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.

Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.  Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.  Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and ACE inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis. The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.  The association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the rate ratios of 2 to 5 for severe COVID-19 associated with designated conditions in this population .

The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019]. Although a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy, this procedure is likely to retain the drug classes that are most commonly prescribed.  We have shown a clear dose-response relationship of dispensed average daily dose of proton pump inhibitors to risk of severe COVID-19.  Adjustment for covariates pre-specified as likely to confound this association -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly. The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.

Where the exposure under study cannot be measured accurately, as in this study that relies on dispensed prescriptions, rather than direct recording of use including over the counter purchasing, adjusting for a propensity score may over-adjust the effect of the exposure under study towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small. One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.  We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.  However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.

It is thus not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.  Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy may be able to resolve this.  As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through effects on the gut microbiome that have been demonstrated [@imhann_proton_2016].  In Scotland and other countries proton pump inhibitors are widely prescribed, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].  It would be of interest to examine the association between severe COVID-19 and use of proton pump inhibitors in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].

We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.  We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020]. The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  We found no evidence that prior use of anticoagulants protects against severed disease, but the association is heavily confounded by factors including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The observed association with hydroxychloroquine is similar to that for other anti-rheumatic drugs, but the numbers exposed to this drug were too small for this association to be estimated reliably.
